Template for indicator documentation

### 2023-12-04 insectIndicators.Rmd
updated to lastest dev

<!--# Notes on this draft
Most things are done but some things needs updates or touching up.

Incomplete list of things that will change:

* Upgrade genetic diversity calculation. The current calculation is just a placeholder, awaiting some new data from Marie. Waiting for software installation by IT. Unsure of timeline.
* Calculate NorIns indicators on 5 year interval. (Moving window?)
* Fix remaining comments in this draft inside hard brackets.-->

@@

This workflow is shortened by putting most of the underlying code in separate R-packages, which are freely available through Github. The relevant packages are `bombLepiSurv` https://github.com/jenast/bombLepiSurv for the bumblebee and butterfly indicators (**Not public yet**), and `Norimon` https://github.com/jenast/Norimon for the rest of the indicators.

@@

```{r, echo=F}
indDataset <- c(rep("NMBB", 2), 
             rep("NorIns", 6))

Indicator  <- c("Bumblebees in semi-natural grasslands", 
                "Butterflies in semi-natural grasslands",
                "Biomass of flying insects",
                "Species richness of flying insects",
                "Species richness of pollinating insects",
                "Species richness of dung associated insects",
                "Quotent of Apocrita vs Symphyta species richness",
                "Intraspesific genetic variation")

nor_cat <- c("Funksjonelt viktige arter og biofysiske stukturer",
             "Funksjonelt viktige arter og biofysiske stukturer",
             "Funksjonelt viktige arter og biofysiske stukturer",
             "Biologisk mangfold",
             "Biologisk mangfold",
             "Biologisk mangfold",
             "Funksjonelt viktige arter og biofysiske strukturer",
             "Biologisk mangfold"
)

ect_cat <- c("B1",
             "B1",
             "B2",
             "B1",
             "B1",
             "B1",
             "B2",
             "B1"
)
Geo_scope <- c(rep("Sørlandet, Østfold-Vestfold, Vestlandet, Trøndelag", 2), 
               rep("Sørlandet, Østlandet, Trøndelag, Nord-Norge", 6))

Ind_data <- c(rep("Present", 2),
              rep("Present", 6))

Ref_val   <- c(rep("Present", 2), 
               rep("Lacking", 6))
```

@@

knitr::kable(indList)
```
Many of these indicators are self-explanatory and don't require much justification. **Bumblebees in semi-natural grasslands** and **Butterflies in semi-natural grasslands** are functionally important pollinators, and are as a community dependent on a varied herbaceous flora. They therefore both represent key functionality, as well as indicators of habitat quality. **Biomass of flying insects** represent a key functional characteristic in ecosystem functioning as food for insectivores, and as a response to habitat quality and management. **Species richness of flying insects** represents the total breadth of the insect community, and **Species richness of pollinating insects** complements the butterfly and bumblebee indicators in taxonomic breadth and spatial resolution.
**Species richness of dung associated insects** represent a key functional group in decomposition, and function as an indicator of grazing intensity and diversity. **Quotent of Apocrita vs Symphyta species richness** tries to go beyond simple amounts, and capture compositional changes in the insect community. Apocrita covers predatory and parasitic wasps, while Symphyta is plant-feeding wasps. A relative decrease in predatory/parasitic wasps could indicate a degradation in insect prey communities, and a decrease in the potential for natural pest control. Lastly, **Intraspecific genetic variation** captures biodiversity that until now has been hidden, due to lack of practical technologies to measure it. The need to conserve genetic diversity is identified in Norwegian law, and decreased intraspecific diversity could signal population decreases, and decreased potential to adapt to environmental changes, before whole species are lost to an area.

## About the underlying data
The data used here comes from the ongoing monitoring programs "Norwegian insect monitoring" ([NorIns](https://www.nina.no/Naturmangfold/Insekter/Overv%C3%A5king-av-insekter)) and "Norwegian monitoring of bumblebees and butterflies" ([NMBB](https://www.nina.no/V%C3%A5re-fagomr%C3%A5der/Milj%C3%B8overv%C3%A5king-p%C3%A5-land/Humler-og-dagsommerfugler)). Both programs are initiated and financed by the Environmental agency, with the aim to produce continuous, areal representative, time-series of insect community data. The **NorIns** program use malaise and window traps to sample a broad community of insect species, which are identified through metabarcoding of DNA. The **NMBB** program use sweep netting along transect walks where the species are identified manually in the field. More information is available in the links above.

@@

These 5 regions represents the planned spatial precision of the survey. Finer resolution could be evaluated in the future, although it is unlikely to be feasible without more intense sampling. This program uses a staggered sampling scheme where we visit different locations each year, and return to each location every 5 years. This sampling scheme, coupled with the strong yearly variation in insect abundances makes it reasonable to use a temporal resolution of 5 years. [How to deal with this now? average over 5 years? Add 5 year aggregation in bootstrap_value()]

@@

Both programs are designed to be area representative, within their respective habitat types and regions. Both use semi-randomized selection schemes of localities, the **NorIns** selects by random draws, and the **NMBB** by a predefined continuous network of survey squares. In both programs however, some subjective decisions are made to define habitat criteria and to discard unpractical survey locations.

@@

## Ecosystem characteristic
### Norwegian standard
Four of the indicators are tied to the *Økologisk egenskap* (Ecological characteristic) of **Funksjonelt viktige arter og biofysiske stukturer** (Funktionally important species and structures), and the other four to **Biologisk mangfold** (Biodiversity).

@@

## Collinearities with other indicators
None measured so far, but it is reasonable to assume that some of these indicators will share some causal relationships with each other, and therefore correlate to some degree. They might also correlate to other indicators, not described here, as well.

@@

### Reference condition
Assessing the current state of insect communities may be a complicated task, due to their taxonomic and functional breadth and high temporal and spatial variability. Still, even more challenging is to establishing reference values, when the reference states are extinct or prohibitively difficult to measure empirically. We currently lack reference values for all indicators from the **NorIns** program. It is unlikely that this can be solved "simply" by surveying a state in a reference area, since intact reference areas of sufficient size and numbers likely no longer exist. The question of how to handle these reference values is currently unresolved, and we present these indicator value calculations with made up reference values as placeholders in the code.

For the **NMBB** indicators, the communities are small enough and well known enough to identify reference communities by expert opinion. These communities specify the expected rate of observing each species. The current rates of observation are then compared to the expected rates. The calculations are further explained below.

### Reference values, thresholds for defining _good ecological condition_, minimum and/or maximum values
<!-- Text here -->
The specific reference communities for the **NMBB** indicators won't be spelled out here, see instead the scripts below. We have not specified a value for _good ecological condition_ as of yet, since there lack a straightforward empirical basis to do so.

## Uncertainties
The uncertainties for the indicator values is calculated by bootstrapping the average values for each locality within a year. This takes the variability between the localities into account. For the **NorIns** indicators, this uncertainties are given as standard deviation, as well as confidence intervals.The uncertainties for the **NMBB** indicators are constituted by a distribution of discrete values, and therefore we only provide confidence intervals and not a standard deviation.

## References
* Åström, Jens; Birkemoe, Tone; Brandsegg, Hege; Dahle, Sondre; Davey, Marie Louise; Ekrem, Torbjørn; Fossøy, Frode; Hanssen, Oddvar; Laugsand, Arne Endre; Majaneva, Markus; Staverløkk, Arnstein; Sverdrup-Thygeson, Anne; Ødegaard, Frode.
Insektovervåking på Østlandet, Sørlandet og i Trøndelag. Rapport fra feltsesong 2022. Trondheim: Norsk institutt for naturforskning (NINA) 2023 (ISBN 978‐82‐426‐5037‐5) 99 s. NINA rapport(2241)

@@

* Connor M French, Laura D Bertola, Ana C Carnaval, Evan P Economo, Jamie M Kass, David J Lohman, Katharine A Marske, Rudolf Meier, Isaac Overcast, Andrew J. Rominger, Phillip Staniczenko, Michael J Hickerson. 2023. Global determinants of insect mitochondrial genetic diversity. https://doi.org/10.1101/2022.02.09.479762

@@

Most of the ecology comes into step 1, in choosing what data to use to describe a quality. This can be a custom selection of species, or a set of higher taxa such as specific genuses, families or even order. The goal is to choose a set of taxa that represent specific qualities of the community that can indicate the ecological status of the ecosystem. In addition to the selection of taxa, we have to decide on the level of aggregation, e.g. if we should aggregate the data on the region scale, locality scale, or even the individual sample occasion scale. For ecological indicators, we will aggregate the values to a year locality scale, where the variables summarises all catches in a locality in a year.[NOTE, implement 5 year/region aggregation]

The second step is to get an estimate of the uncertainty of the data, were we use the bootstrap method. This method is flexible, as we could bootstrap the samples on different sampling levels (samples within localities, regions within country). For the insect indicators, we will bootstrap and summarize the variables to regional values, for example expressing the mean values of insect biomass in Trøndelag in 2021, with uncertainty.

@@

Taken together, this framework is meant to facilitate the calculation of an arbitrary set of insect indicators, based on the combination of choices in data to fetch, aggregation level, and reference comparison. We can work through an example with pollinators, to showcase the framework.

In the example, we start with connecting to the database in order to fetch the data.

@@

```{r, echo = FALSE}
kable(pollinators)
```

We then fetch the community data for these families through the `get_observations` function. We additionally subset the habitat type that we are interested in, here semi-natural land. The result is a tibble of the aggregated number of species, shannon diversity, and the mean number of genetic variants per species. The default aggregation level is "year_locality", meaning the total observations for a locality within a year.

Note that the output currently contains some experimental values, awaiting a more robust methodology. "Shannon diversity" doesn't really make sense without counts or amounts, etc. This will be replaced by measurements of genetic diversity that are based on published peer-reviewed methodology (Connor et al. 2023), in the (hopefully near) future.

@@

#### Bootstrap observations
The localities in the NorIns program are (semi-)randomly selected and can be viewed as samples of a larger population. Individually, they represent a single measurement of an insect community in a habitat type in a region. To get the sampling uncertainty for this representation, we can bootstrap the values by this simple process: we choose a random set of localities within a year and region (with replacement) and calculate the average values. We then repeat this process a large number of times to get a bootstrap sample of values, which can be used to express the uncertainty in the dataset.

We typically have 10 localities for a given habitat type and region each year, which is not very much to base our uncertainty estimates on. But bootstrapping works fairly well on small samples and has the advantage that is doesn't make assumptions of the statistical distribution of the errors. Note that once we get a larger dataset with repeating surveys every 5 years, aggregating and bootstrapping over a whole 5 year period.

@@

poll_richness_boot <- bootstrap_value(poll_loc_year,
                                      value = no_species,
                                      groups = c("year",
                                                 "region_name")
                                      )

@@

#### Comparing bootstrap values to a reference point
Setting aside the practical difficulties in establishing robust reference values, the next step in the methodology is to compare the observed values (with bootstrapped uncertainty) to a chosen reference value. This can be made in several ways. Most simply, if we have a single numeric value as a reference value, we can simply subtract that from the observed values. For example, if we observe 13 species of pollinators in 2022 at a location, and the reference point is 10, the 2022 value has increased by 13 - 10 = 3 species. Such subtractions should be made on the set of bootstrapped values, followed by new summary statistics being calculated, preserving the uncertainty from the bootstrap. The `boot_stat` has its own subtraction method `-` do to just that. In this example, we set the reference value arbitrarily to 30 species.

@@

#### Normalizing the values
After the comparison to a reference value, we need to normalize the indicator values so that they lie between 0 and 1. This can be done in several ways (se e.g. [eaTools](https://github.com/NINAnor/eaTools)). The simplest case is to use a linear scaling, with a natural zero, which e.g. can be done by dividing the indicator values by the highest value state. A `boot_stat` class has a `/` function that divides each bootstrap value by a given value, truncates the highest values to 1, and recalculates the summary values. We can also simply divide by the reference state, to compare the values to the reference state, and normalize them in one go. Here, we exemplify the method by dividing the value by 30.

@@

@@

#### Map plots
In addition to the density plots above, we can also display the values geographically. The `map_plot()` function takes a `boot_stat` object and plots the values according to its region names.

By default, it only shows the regions with data, but this can be overridden manually:
```{r map_plot_whole}
map_plot(poll_richness_boot,
         whole_country = TRUE)
```

We can also choose a different palette, for example from the NinaR package, and visualize the uncertainty by setting the transparency of the colors from the bootstrap standard deviations. By the way, most of these functions can be piped as well:

@@

$$CI = \frac{SR-SC}{SR}$$
The state of reference (SR) represents a community of species that can be expected to be observed in a given habitat type and region. It is calculated by assigning each species to class of expected commonality: common (C), sporadic (S) and rare (R) species. This classification is done by expert opinion, informed by known present and past species distributions. Note that this state of reference only contains a subset of highlighted species, making up a historical reference community. Potential observations of "new" species therefore don't inform the indicator value at all. Common species are expected to be observed in at least 5 % of the surveyed transects in a habitat type and region. Sporadic species similarly are defined as having a presence below 5 % but above 1 %. Rare species are seen in no more than 1 % of the transects, and lastly, species not seen in any transect are assumed as lost (L) for the purpose of this calculation.

@@

The state of change (SC) is calculated as
 $$SC = n_{CS} * w_{CS} + n_{CR} * w_{CR} + n_{CL} * w_{CL} + n_{SR} * w_{SR} + n_{SL} * w_{SL} + n_{RL} * w_{RL}, $$ 
where $n_{CS}$ is the number of expected common species (C) that is observed sporadically (S), and $w_{CS}$ is the weight of this change. Similarly, $n_{CR}$ is the number of expected species that is observed rarely, $n_{RL}$ is the number of rare species that is lost, and so on. Potential increases in observation rates do not inform the indicator. Changes for common species are weighted more heavily than changes for less common species, and larger decreases are weighted more heavily than smaller decreases, such that $[w_{CS}, w_{CR}, w_{CL}, w_{SR}, w_{SL}, w_{RL}]= [0.50, 0.75, 1.00, 0.50, 0.75, 0.50]$.

@@

Plotting functions for a series of indicator values is shown below in the indicator calculations.

@@

#### NorIns data
The `Norimon` package has convenience functions to fetch data from the database, housed internally at NINA. This database is currently not available outside NINA, but we will implement a solution for this. Either we will make the database externally available, or we will provide an alternative route to the data from the GBIF export. For now, we will fetch the data through the `Norimon` functions.

#### NMBB data
The data for the bumblebee and butterfly indicators can be accessed through the R-package `bombLepiSurv`.

@@

The indicators from the NMBB program are similary connected to regions, with the exception that the south-east region only covers the old counties of Vestfold and Østfold, and that Nord-Norge isn't covered yet.

@@

#### Biomass of flying insects in semi-natural land

Fetch the data.
```{r}
biomass_sn <- get_biomass(subset_year = 2020:2022,
                          subset_region = NULL)
```

@@

Compare values to a reference state. The reference state is here set uniformly for all regions, with a made-up value just to be able to continue the calculations. Alternatively, we could calculate the indicator values for each region separately, compare to individual reference states, and then put it all together again.

We here use a single value as a reference state and normalize the values at the same time, using the `/` function.

@@

```{r}
plot(biomass_sn_diff)
```

@@

```{r}
biomass_sn_to_exp <- biomass_sn_diff$bootstrap_summary

biomass_sn_to_exp <- biomass_sn_to_exp %>% 
  select(year,
         region = region_name,
         ind_value = boot_value,
         ind_sd = boot_sd,
         ind_lower2.5 = boot_lower2.5,
         ind_upper97.5 = boot_upper97.5)

@@

biomass_sn_to_exp
```

@@

```{r}
richness_sn <- get_observations(subset_year = 2020:2022,
                                subset_region = NULL)
```

@@

```{r}
plot(richness_sn_diff)
```

@@

```{r}
richness_sn_to_exp <- richness_sn_diff$bootstrap_summary

richness_sn_to_exp <- richness_sn_to_exp %>% 
  select(year,
         region = region_name,
         ind_value = boot_value,
         ind_sd = boot_sd,
         ind_lower2.5 = boot_lower2.5,
         ind_upper97.5 = boot_upper97.5)

@@

richness_sn_to_exp
```

@@

```{r}
pollinators_sn <- get_observations(subset_year = 2020:2022,
                                   subset_families = pollinator_fam,
                                   subset_region = NULL)
```

@@

```{r}
plot(pollinators_sn_diff)
```

@@

```{r}
pollinators_sn_to_exp <- pollinators_sn_diff$bootstrap_summary

pollinators_sn_to_exp <- pollinators_sn_to_exp %>% 
  select(year,
         region = region_name,
         ind_value = boot_value,
         ind_sd = boot_sd,
         ind_lower2.5 = boot_lower2.5,
         ind_upper97.5 = boot_upper97.5)

@@

pollinators_sn_to_exp
```

#### Species richness of dung associated insects
Note that the species list will probably be expanded, after a taxonomic review.

>>>>

@@

This workflow is shortened by putting most of the underlying code in separate R-packages, which are freely available through Github. The relevant packages are `bombLepiSurv` https://github.com/jenast/bombLepiSurv for the bumblebee and butterfly indicators, and `Norimon` https://github.com/jenast/Norimon for the rest of the indicators.

@@

```{r, echo=F}
indDataset <- c(rep("NMBB", 2), 
             rep("NorIns", 7))

Indicator  <- c("Bumblebees in semi-natural grasslands", 
                "Butterflies in semi-natural grasslands",
                "Biomass of flying insects",
                "Species richness of flying insects",
                "Species richness of pollinating insects",
                "Species richness of dung associated insects",
                "Quotent of Apocrita vs Symphyta species richness",
                "Number of alien species",
                "Intraspesific genetic variation")

nor_cat <- c("Funksjonelt viktige arter og biofysiske stukturer",
             "Biologisk mangfold",
             "Funksjonelt viktige arter og biofysiske stukturer",
             "Biologisk mangfold",
             "Biologisk mangfold",
             "Biologisk mangfold",
             "Funksjonelt viktige arter og biofysiske strukturer",
             "Biologisk mangfold",
             "Biologisk mangfold"
)

ect_cat <- c("B2",
             "B1",
             "B2",
             "B1",
             "B1",
             "B1",
             "B2",
             "B1",
             "B1"
)
Geo_scope <- c(rep("Sørlandet, Østfold-Vestfold, Vestlandet, Trøndelag", 2), 
               rep("Sørlandet, Østlandet, Trøndelag, Nord-Norge", 7))

Ind_data <- c(rep("Present", 2),
              rep("Present", 7))

Ref_val   <- c(rep("Present", 2), 
               rep("Lacking", 7))
```

@@

knitr::kable(indList)
```
Many of these indicators are self-explanatory and don't require much justification. **Bumblebees in semi-natural grasslands** and **Butterflies in semi-natural grasslands** are functionally important pollinators, and are as a community dependent on a varied herbaceous flora. They therefore both represent key functionality, as well as indicators of habitat quality. **Biomass of flying insects** represent a key functional characteristic in ecosystem functioning as food for insectivores, and as a response to habitat quality and management. **Species richness of flying insects** represents the total breadth of the insect community, and **Species richness of pollinating insects** complements the butterfly and bumblebee indicators in taxonomic breadth and spatial resolution.
**Species richness of dung associated insects** represent a key functional group in decomposition, and function as an indicator of grazing intensity and diversity. **Quotent of Apocrita vs Symphyta species richness** tries to go beyond simple amounts, and capture compositional changes in the insect community. Apocrita covers predatory and parasitic wasps, while Symphyta is plant-feeding wasps. A relative decrease in predatory/parasitic wasps could indicate a degradation in insect prey communities, and a decrease in the potential for natural pest control. This indicator is optionally given as just the species richness of Apocrita. **Number of alien species** takes into account the number of alien species found in a region and optionally, the frequency of sampling events they are observed in. This is based on the alien species list from the Norwegian Biodiversity Information Center, which lists known alien species and an increasing list of door-knocker species. Note that due to large historical knowledge gaps of insect species distributions, there could potentially be many more alien species in Norway, which are not picked up on this alien species list. Lastly, **Intraspecific genetic variation** captures biodiversity that until now has been hidden, due to lack of practical technologies to measure it. The need to conserve genetic diversity is identified in Norwegian law, and decreases in intraspecific diversity could signal population decreases, and a decreased potential to adapt to environmental changes, before whole species are lost to an area.

## About the underlying data
The data used here comes from the ongoing monitoring programs "Norwegian insect monitoring" ([NorIns](https://www.nina.no/Naturmangfold/Insekter/Overv%C3%A5king-av-insekter)) and "Norwegian monitoring of bumblebees and butterflies" ([NMBB](https://www.nina.no/V%C3%A5re-fagomr%C3%A5der/Milj%C3%B8overv%C3%A5king-p%C3%A5-land/Humler-og-dagsommerfugler)). Both programs are initiated and financed by the Environmental agency, with the aim to produce continuous area representative time-series of insect community data. The **NorIns** program use malaise and window traps to sample a broad community of insect species, which are identified through metabarcoding of DNA. The **NMBB** program use sweep netting along transect walks where the species are identified manually in the field. More information is available in the links above.

@@

The monitoring program is designed to produce estimates with a spatial resolution of these 5 large regions. Finer resolution could be evaluated in the future, although it is unlikely to be feasible without more intense sampling. The program uses a staggered sampling scheme where we visit different locations each year, and return to each location every 5 years. This sampling scheme, coupled with the strong yearly variation in insect abundances makes it reasonable to use a temporal resolution of 5 years. The indicator values from **NorIns** are therefore calculated on 5 year long rolling windows.

@@

Both programs are designed to be area representative, within their respective habitat types and regions. Both use stratified randomized selection of localities. The **NorIns** program selects localities by random draws within habitat strata, and the **NMBB** by a predefined continuous network of survey squares, within which individual transects are then manually placed. In both programs, some subjective decisions are made in the definition of habitat criteria and in discarding unpractical survey locations.

@@

## Ecosystem characteristic
### Norwegian standard
Three of the indicators are tied to the *Økologisk egenskap* (Ecological characteristic) of **Funksjonelt viktige arter og biofysiske stukturer** (Funktionally important species and structures), and the other six to **Biologisk mangfold** (Biodiversity).

@@

## Collinearities with other indicators
None measured so far, but it is reasonable to assume that some of these indicators will share some causal factors, and therefore correlate to some degree. They might also correlate to other indicators as well, not described here.

@@

### Reference condition
Assessing the current state of insect communities is a complicated task, due to their taxonomic and functional breadth and high temporal and spatial variability. Still, even more challenging is establishing reference values, when the reference states are extinct or prohibitively difficult to measure empirically. We currently lack reference values for all indicators from the **NorIns** program, except for the alien species indicators where the reference value is zero by definition. It is unlikely that this can be solved by "simply" surveying a state in a reference area, since intact reference areas of sufficient size and numbers likely no longer exist. The question of how to handle these reference values is currently unresolved, and we present these indicator value calculations with made up reference values as placeholders in the code.

For the **NMBB** indicators, the communities are small enough and well-known enough to identify reference communities by expert opinion. These communities specify the expected rate of observing each species. The current rates of observation are then compared to the expected rates. The calculations are further explained below.

### Reference values, thresholds for defining _good ecological condition_, minimum and/or maximum values
<!-- Text here -->
The specific reference communities for the **NMBB** indicators won't be spelled out here, see instead the scripts below. We have not specified a value for _good ecological condition_ as of yet, since we lack a straightforward empirical basis to do so. By default then, the values is set to 0.6 , following the general framework.

## Uncertainties
The uncertainties for the indicator values is calculated by bootstrapping the values for each locality within a year. This takes the variability between the localities into account. For the **NorIns** indicators, these uncertainties are given as standard deviation, as well as confidence intervals. The uncertainties for the **NMBB** indicators are constituted by a distribution of discrete values, and therefore we only provide confidence intervals and not a standard deviation.

## References

* French, C. M., Bertola, L. D., Carnaval, A. C., Economo, E. P., Kass, J. M., Lohman, D. J., … Hickerson, M. J. 2023, juli 17. Global determinants of insect mitochondrial genetic diversity. bioRxiv. bioRxiv. https://doi.org/10.1101/2022.02.09.479762

* Miraldo, A., Li, S., Borregaard, M. K., Flórez-Rodríguez, A., Gopalakrishnan, S., Rizvanovic, M., … Nogués-Bravo, D. 2016. An Anthropocene map of genetic diversity. Science 353(6307): 1532–1535. https://doi.org/10.1126/science.aaf4381

* Rohde, K. 1992. Latitudinal Gradients in Species Diversity: The Search for the Primary Cause. Oikos 65(3): 514. https://doi.org/10.2307/3545569

* Theodoridis, S., Fordham, D. A., Brown, S. C., Li, S., Rahbek, C., & Nogues-Bravo, D. 2020. Evolutionary history and past climate change shape the distribution of genetic diversity in terrestrial mammals. Nature Communications 11(1): 2557. https://doi.org/10.1038/s41467-020-16449-5

* Åström, Jens; Birkemoe, Tone; Brandsegg, Hege; Dahle, Sondre; Davey, Marie Louise; Ekrem, Torbjørn; Fossøy, Frode; Hanssen, Oddvar; Laugsand, Arne Endre; Majaneva, Markus; Staverløkk, Arnstein; Sverdrup-Thygeson, Anne; Ødegaard, Frode.
Insektovervåking på Østlandet, Sørlandet og i Trøndelag. Rapport fra feltsesong 2022. Trondheim: Norsk institutt for naturforskning (NINA) 2023 (ISBN 978‐82‐426‐5037‐5) 99 s. NINA rapport(2241)

@@

@@

Most of the ecology comes into step 1, in choosing what data to use to describe a quality. This can be a custom selection of species, or a set of higher taxa such as specific genuses, families or even order. The goal is to choose a set of taxa that represent specific qualities of the community that can indicate the ecological status of the ecosystem. In addition to the selection of taxa, we have to decide on the level of spatial and temporal aggregation, e.g. if we should aggregate the data spatially on the region scale, habitat scale or locality scale, and if we should aggregate it temporally on yearly or even the individual sample occasions within years. For the ecological indicators, we will aggregate the raw observation values to the year locality scale, where the data summarize all catches in a locality in a year.

The second step is to get an estimate of the uncertainty of the data, were we use the bootstrap method. This method is also flexible, as we could bootstrap the samples on different sampling levels (e.g. samples within localities, localities within a region, or regions within country). When working with raw data on the year locality level, the most granular bootstrap would be to summarize the variables to the yearly, habitat, and regional level, for example expressing the mean values of insect biomass in semi-natural land and forests in Østlandet in 2021, with uncertainty.

@@

Taken together, this framework is meant to facilitate the calculation of an arbitrary set of insect indicators, based on the combination of choices in data to fetch, aggregation level, and reference comparison. We showcase the framework by working through an example with pollinators.

In the example we start with connecting to the database, in order to fetch the data.

@@

```{r, echo = FALSE}
kable(pollinators)
```
<br/>
We then fetch the community data for these families through the `get_observations` function. We additionally subset the habitat type that we are interested in, here semi-natural land. The result is a tibble of the aggregated number of species, shannon diversity, and the mean number of genetic variants per species. The default aggregation level is "year_locality", meaning the total observations for a locality within a year.

Note that the output currently contains some experimental values, awaiting a more robust methodology. "Shannon diversity" doesn't really make sense without counts or amounts, etc. This will be replaced by measurements of genetic diversity that are based on published peer-reviewed methodology (French et al. 2023), in the (hopefully near) future.

@@

#### Bootstrap observations
The localities in the NorIns program are (semi-)randomly selected and can be viewed as samples of a larger population. Individually, they represent a single measurement of an insect community in a habitat type in a region. To get the sampling uncertainty for this representation, we can bootstrap the values by this simple process: we choose a random set of localities within a year and region (with replacement) and calculate the average value. We then repeat this process a large number of times to get a bootstrap sample of values, which can be used to express the uncertainty in the dataset.

We typically have 10 localities for a given habitat type and region each year, which is not very much to base our uncertainty estimates on. But bootstrapping works fairly well on small samples and has the advantage that is doesn't make assumptions of the statistical distribution of the errors. We will use this level to showcase the functionality and to visualize some variation between years. However, since the sampling scheme of the monitoring program semi-randomly selects 10 new localities for each of the 5 years the rotating scheme, each years estimate will include some randomness that is caused by variability between localities. This means that the values between individual years might appear more random than the underlying overall trend. To get robust estimates that averages over this variability, we will summarize the actual indicator values to 5 year periods, further described below.

@@

poll_richness_boot <- bootstrap_value(poll_loc_year,
                                      value = no_species,
                                      groups = c("year",
                                                 "region_name"),
                                      rolling_year_window = FALSE
                                      )

@@

#### Comparing bootstrap values to a reference point
Setting aside the practical difficulties in establishing robust reference values, the next step in the methodology is to compare the observed values (with bootstrapped uncertainty) to a chosen reference value. This can be made in several ways. Most simply, if we have a single numeric value as a reference value, we can simply subtract that from the observed values. For example, if we observe 13 species of pollinators in 2022 at a location, and the reference point is 10, the 2022 value has increased by 13 - 10 = 3 species. Such subtractions should be made on the set of bootstrapped values, followed by new summary statistics being calculated, preserving the uncertainty from the bootstrap. The `boot_stat` object class has its own subtraction method `-` do to just that. In this example, we set the reference value arbitrarily to 30 species.

@@

#### Normalizing the values
After the comparison to a reference value, we need to normalize the indicator values so that they lie between 0 and 1. This can be done in several ways (se e.g. [eaTools](https://github.com/NINAnor/eaTools)). The simplest case is to use a linear scaling, with a natural zero, which e.g. can be done by dividing the indicator values by the highest value state. A `boot_stat` class has a `/` function that divides each bootstrap value by a given value, truncates the highest values to 1, and recalculates the summary values. We can similarly divide by a predefined reference state. Here, we exemplify the method by dividing the value by 30.

@@

#### Time-series plots
The default plots are good for showing the uncertainty distributions, but can be a bit difficult to track the time series trends this way. We can instad use the `ts_plot` function for a simple time-series plot.

```{r}
ts_plot(poll_richness_boot)
```

#### Map plots
In addition to these plots, we can also display the values geographically. The `map_plot()` function takes a `boot_stat` object and plots the values according to its region names.

By default, it only shows the regions with data, but this can be overridden manually:
```{r map_plot_whole}
map_plot(poll_richness_boot,
         whole_country = TRUE)
```
We can also choose a different palette, for example from the NinaR package, and visualize the uncertainty by setting the transparency of the colors from the bootstrap standard deviations. Most of these functions can be piped as well:

@@

#### 5 year rolling window summaries
As described above, to take the staggered 5 year survey scheme into account and to average over random yearly variation, it makes sense to summarize the state over a whole 5 year period. This does not need to adhere to predefined periods, since any 5 year long period will result in a complete data set. We can therefore calculate 5 year long summaries in rolling windows, where each focal year is surrounded by +- 2 years. The bootstrap_value does this by default, if we don't specify `rolling_year_average = FALSE`

```{r}
poll_richness_boot_5_year <- bootstrap_value(poll_loc_year,
                                      value = no_species,
                                      groups = c("year",
                                                 "region_name"),
                                      rolling_year_window = TRUE
                                      )
```
Since we are only on the third year in the survey scheme however, the rolling windows are too large to show any differences between years. This will change, starting with the 2023 year data.

```{r}
plot(poll_richness_boot_5_year)
```

The averaging over 5-years also shows up in the time-series plots. Note the very wide confidence intervals of the Sørlandet region, with only 1-year worth of data.

```{r}
ts_plot(poll_richness_boot_5_year)
```

@@

$$CI = \frac{SR-SC}{SR}$$
The state of reference (SR) represents a community of species which can be expected to be observed in a given habitat type and region. It is calculated by assigning each species to class of expected commonality: common (C), sporadic (S) and rare (R) species. This classification is done by expert opinion, informed by known present and past species distributions. Note that this state of reference only contains a subset of highlighted species, making up a historical reference community. Potential observations of "new" species therefore don't inform the indicator value at all. Common species are expected to be observed in at least 5 % of the surveyed transects in a habitat type and region. Sporadic species similarly are defined as having a presence below 5 % but above 1 %. Rare species are seen in no more than 1 % of the transects, and lastly, species not seen in any transect are assumed as lost (L) for the purpose of this calculation.

@@

The state of change (SC) is calculated as
 $$SC = n_{CS} * w_{CS} + n_{CR} * w_{CR} + n_{CL} * w_{CL} + n_{SR} * w_{SR} + n_{SL} * w_{SL} + n_{RL} * w_{RL}, $$ 
where $n_{CS}$ is the number of expected common species (C) that is observed sporadically (S), and $w_{CS}$ is the weight of this change. Similarly, $n_{CR}$ is the number of expected common species that is observed rarely, $n_{RL}$ is the number of rare species that is lost, and so on. Potential increases in observation rates do not inform the indicator. Changes for common species are weighted more heavily than changes for less common species, and larger decreases are weighted more heavily than smaller decreases, such that $[w_{CS}, w_{CR}, w_{CL}, w_{SR}, w_{SL}, w_{RL}]= [0.50, 0.75, 1.00, 0.50, 0.75, 0.50]$.

@@

Plotting functions for a series of indicator values is shown below in the indicator calculations. This concludes the tour of the `bombLepiSurv` package.

@@

#### NorIns data
The `Norimon` package has convenience functions to fetch data from the database. This database is currently not available outside NINA, but we will implement a solution for this in the future. Either we will make the database externally available, or we will provide an alternative route to the data from the GBIF export. For now, we will fetch the data through the `Norimon` functions.

#### NMBB data
The data for the bumblebee and butterfly indicators can similarly be accessed through the R-package `bombLepiSurv`.

@@

The indicators from the NMBB program are similarly connected to regions, with the exception that the south-east region only covers the old counties of Vestfold and Østfold, and that Nord-Norge isn't covered yet.

```{r, eval=FALSE}
bombLepiSurv::humlesommerfConnect()
```

@@

#### Biomass of flying insects in semi-natural land
```{r, eval=FALSE}
connect_to_insect_db()
```

Fetch the data.
```{r}
biomass_sn <- get_biomass(subset_year = 2020:2022,
                          subset_region = NULL,
                          subset_habitat = "Semi-nat")
```

@@

Compare values to a reference state. The reference state is here set uniformly for all regions. Alternatively, we could calculate the indicator values for each region separately, compare to individual reference states, and then put it all together again.

To exemplify, we here use a single "made-up" value as a reference state and normalize the values at the same time, using the `/` function.

@@

```{r biomass_diff}
plot(biomass_sn_diff)
```

@@

```{r}
biomass_sn_to_exp <- biomass_sn_boot$bootstrap_summary

biomass_sn_to_exp <- biomass_sn_to_exp %>% 
  mutate(reference_high = NA,
         reference_low = 0,
         thr = 0.6,
         i = NA) %>% 
  select(year,
         region = region_name,
         v = boot_value,
         sd = boot_sd,
         reference_high,
         reference_low,
         thr,
         i
         )

@@

insect_biomass_semi_nat <- biomass_sn_to_exp
```

@@

```{r}
richness_sn <- get_observations(subset_year = 2020:2022,
                                subset_region = NULL,
                                subset_habitat = "Semi-nat")
```

@@

```{r}
plot(richness_sn_diff,
     limit = c(0, 1))
```

```{r richness}
map_plot(richness_sn_diff,
         whole_country = T,
         limit = c(0, 1))
```

@@

```{r}
richness_sn_to_exp <- richness_sn_boot$bootstrap_summary

richness_sn_to_exp <- richness_sn_to_exp %>% 
  mutate(reference_high = NA,
         reference_low = 0,
         thr = 0.6,
         i = NA) %>% 
  select(year,
         region = region_name,
         v = boot_value,
         sd = boot_sd,
         reference_high,
         reference_low,
         thr,
         i
         )

@@

insect_richness_semi_nat <- richness_sn_to_exp
```

@@

```{r}
pollinators_sn <- get_observations(subset_year = 2020:2022,
                                   subset_families = pollinator_fam,
                                   subset_habitat = "Semi-nat")
```

@@

```{r}
plot(pollinators_sn_diff,
     limit = c(0, 1))
```

@@

```{r}
pollinators_sn_to_exp <- pollinators_sn_boot$bootstrap_summary

pollinators_sn_to_exp <- pollinators_sn_to_exp %>% 
  mutate(reference_high = NA,
         reference_low = 0,
         thr = 0.6,
         i = NA) %>% 
  select(year,
         region = region_name,
         v = boot_value,
         sd = boot_sd,
         reference_high,
         reference_low,
         thr,
         i
         )

@@

insect_pollinators_semi_nat <- pollinators_sn_to_exp
```
